---
title: "RMS Titanic"
author: "Lee Edwards"
date: 2020-07-15
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
glimpse(df_titanic)
```

**Observations**:

Class, Sex, Age, Survived, n

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

Wikipedia lists two possible numbers for the number of people onboard the Titanic: "3,327 (or 3,547 according to other sources)," but crew specifically as 2,435. The fact that there are conflicting figures for the number of total people on the ship suggests there is some conflicting information in the historical record that might account for the disparity between passenger compliment. Maybe some counts included stowaways, or others included people who bought a ticket but didn't show up.

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}

df_titanic %>%
  filter(Survived == "Yes") %>%
  count(Sex, Class) %>%
  
  ggplot() + 
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge")

```

**Observations**:

Total number of survivors among First and Second Classes seem to be skewed toward women, where among Third Class, it is more equal. The total number of survivors in First is greater than in Second, which is greater than in Third. A much larger numbe rof crew was aved overall, but mostly men, and few women. But we don't know much about the relative sizes of these categories.


# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%

  ggplot() + 
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")
```

**Observations**:

Proportionally, it looks like more First and Second Class passengers survived. In Third Class and Crew, proportionally far did. Among those two groups, more women were saved than men (proportionally.)

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%

  ggplot() + 
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  facet_wrap(~ Age)
```

**Observations**:

Taking children into account, it's clear that many of the surviving men were children. It still seems to be true that lower classes made up a small proportion of survivors. Sadly, this was even true for children, and for women.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
